---
title: "2: Dimension Reduction"
params: 
    SHOW_SOLS: FALSE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(xaringanExtra)
library(tidyverse)
library(patchwork)
library(ggdist)
xaringanExtra::use_panelset()
qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
library(lme4)
```

:::lo

This reading:

- build the intuition for thinking about "dimensions" of data
- introduce some key goals and methods that researchers use, and how these differ in relation to finding and using dimensions of data


:::

# The dimensionality of data

Take a moment to think about the various things that you are often interested in when discussing psychology. This might be anything from personality traits, to language proficiency, social identity, anxiety etc. 
How we measure such constructs is a very important consideration for research. The things we're interested in are very rarely the things we are *directly* measuring. 

Consider how we might assess levels of anxiety or depression. Can we ever directly measure anxiety?^[Even if we cut open someone's brain, it's unclear what we would be looking for in order to 'measure' it. It is unclear whether anxiety even exists as a physical thing, or rather if it is simply the overarching concept we apply to a set of behaviours and feelings.] More often than not, we measure these things using a _set_ of different measures which 'look at' the underlying construct from a different angle. In psychology, this is often questionnaire based methods, with a set of questions each of which might ask about "anxiety" from a slightly different angle. Twenty questions all measuring different aspects of anxiety are (we hope) going to correlate with one another if they are capturing some commonality (the construct of "anxiety").  

But this introduces a problem for us, which is how to deal with 20 variables that represent (in broad terms) the same single thing. How can we talk about "changes in anxiety", and not just "changes on anxiety Q1" and "changes on anxiety Q2" and so on? In addition, not all of these constructs might have just one single dimension - we often talk about "narcissm", but this could arguably be comprised of 3 slightly distinct constructs of "grandiosity", "vulnerability" and "antagonism".  

What we are touching upon here gets referred to as *dimensionality* of a measurement tool, and the questions we are going to look at are all about how we can go from the big set of observed variables to the key *dimensions* that are the things we are actually interested in.  

:::sticky
__When might we want to reduce dimensionality?__

There are many reasons we might want to "reduce the dimensionality" of data:  
- Pragmatic
  - I have multicollinearity issues/too many variables, how can I defensibly combine my variables? 
- Test construction
  - How should I group my questionnaire items into sub-scales?
  - Which items are the best measures of my construct(s)?
- Theory testing
  - What are the number and nature of dimensions that best describe a theoretical construct?
  
:::


# The core idea


```{r include=FALSE}
library(rgl)
library(psych)
set.seed(33)
S = runif(3,.4,2)
f = runif(3,.7,.99)
R = f %*% t(f)
diag(R) = 1
R[1,3] <- R[2,3] <- R[3,1] <- R[3,2] <- .1
Sigma = diag(S)%*%R%*%diag(S)
Mean <- rep(0,3)
x <- MASS::mvrnorm(500, Mean, Sigma)
mydata <- as.data.frame(x)
names(mydata) <- c("y1","y2","y3")
```

Imagine we had 3 measured variables: y1, y2, and y3, as visualised in 3-dimensional space in @fig-scat.   

::::{.columns}
:::{.column width="50%"}
```{r}
#| echo: false
#| label: fig-scat
#| fig-cap: "3 measured variables"
plot3d(x, box = FALSE, xlab="y1",ylab="y2",zlab="y3")
rglwidget()
```

:::
:::{.column width="50%"}
```{r}
#| echo: false
#| label: fig-ellip
#| fig-cap: "An ellipse capturing the cloud of datapoints, the axes shows the length, width and depth of the ellipse"
plot3d(x, box = FALSE, xlab="y1",ylab="y2",zlab="y3")
plot3d(ellipse3d(Sigma, centre = Mean), col = "#A41AE4", alpha = 0.4, add = TRUE)
plot3d(
  abclines3d(0,0,0,a=principal(x,nfactors=3,rotate="none")$loadings[,1],
             col="green"),lwd=2, add=TRUE)
plot3d(
  abclines3d(0,0,0,a=principal(x,nfactors=3,rotate="none")$loadings[,2],
             col="red"), lwd=2, add=TRUE)
plot3d(
  abclines3d(0,0,0,a=principal(x,nfactors=3,rotate="none")$loadings[,3],
           col="blue"), lwd=2, add=TRUE)
rglwidget()

```
:::
::::

Although we have measured 3 variables (`y1`, `y2`, `y3`), when faced with trying to characterise the shape of the 3-dimensional cloud of datapoints, we're more inclined to think about its length, width and depth. You can imagine trying to characterise this shape as the ellipse in @fig-ellip. It looks a little like a bar of soap or something! It's longer in one direction (green line), wide in another direction (red line), and not very deep (blue line). We can think of these as the dimensions that characterise the data.  

::: {.callout-note collapse="false"}
#### a correlation matrix

The example here has 3 variables, which means 3-dimensions, and this is why we can actually visualise it. But we could also think of this shape as being described by all the correlations between each pair of variables. Each of these correlations is characterising the plot of each pair of variables, and these are just like viewing the 3-d plot above but looking at it from straight on at each side.^[Rotate the 3-d plot to make it match the 2-d plots below!] So in cases where we have lots and lots of variables, we might not be able to visualise the n-dimensional shape itself, but we can look at it from all different angles through the correlation matrix!  

::::{.columns}
:::{.column width="48%"}
```{r}
cor(mydata)
```
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
```{r}
pairs(mydata)
```
:::
::::

:::

Think a little about what these dimensions (the axes in @fig-ellip) *represent* - they capture "the ways in which observations vary" across a set of observed variables. The long dimension captures that `y1` and `y2` tend to co-vary (there are lots of people who are high on both `y1` and `y2`, and lots who are low on both). The wider dimension captures that there's another source of variability, mainly tied to being high or low on `y3`. 

Often it helps to take this to the extremes, so think about what these would look like if we *no correlations* or *perfect correlations* between the three variables. If the variables are uncorrelated, then we have a sort of spherical cloud of data, and no one dimension is longer than the other (@fig-pcano). By contrast, if we have perfect correlations between the variables, then we only really have *one* dimension - we just have a line upon which all observations fall (@fig-pcaff).  

::::{.columns}
:::{.column width="50%"}
```{r}
#| echo: false
#| label: fig-pcano
#| fig-cap: "Principal components for 3 uncorrelated variables"
set.seed(33)
S = runif(3,.4,2)
f = runif(3,0,.1)
R = f %*% t(f)
diag(R) = 1
Sigma2 = diag(S)%*%R%*%diag(S)
Mean2 <- rep(0,3)
x2 <- MASS::mvrnorm(500, Mean2, Sigma2)
plot3d(x2, box = FALSE, xlab="y1",ylab="y2",zlab="y3")
plot3d( ellipse3d(Sigma2, centre = Mean2), col = "#A41AE4", alpha = 0.4, add = TRUE)
plot3d(
  abclines3d(0,0,0,a=principal(x,nfactors=3,rotate="none")$loadings[,1],
             col="green"),lwd=2, add=TRUE)
plot3d(
  abclines3d(0,0,0,a=principal(x,nfactors=3,rotate="none")$loadings[,2],
             col="red"), lwd=2, add=TRUE)
plot3d(
  abclines3d(0,0,0,a=principal(x,nfactors=3,rotate="none")$loadings[,3],
           col="blue"), lwd=2, add=TRUE)
rglwidget()
```
:::
:::{.column width="50%"}

```{r}
#| echo: false
#| label: fig-pcaff
#| fig-cap: "Principal components for 3 perfectly correlated variables"
S = runif(3,.4,2)
f = runif(3,1,1)
R = f %*% t(f)
diag(R) = 1
Sigma2 = diag(S)%*%R%*%diag(S)
Mean2 <- rep(0,3)
x2 <- MASS::mvrnorm(500, Mean2, Sigma2)

plot3d(x2, box = FALSE, xlab="y1",ylab="y2",zlab="y3")
rglwidget()
```
:::
::::


Imagine being given a ruler and being asked to give two numbers to provide a measurement of your smartphone. What do you pick? Chances are, you will measure its length and then its width. You're likely to ignore the depth because it is much smaller than the other two dimensions. This is pretty much the concept of dimension reduction. 

The idea here is that we can --- without losing too much --- preserve the general shape of our data using _fewer_ dimensions. Ignoring the depth of our cloud of datapoints, we could reduce from 3 variables down to 2 dimensions by considering where our data falls when projected onto the 2-dimensional surface shown in @fig-ellipreduc. Or we could reduce down to 1 dimension and consider where the data falls on the single axis shown in @fig-ellipreduc2.  

::::{.columns}
:::{.column width="50%"}
```{r}
#| echo: false
#| label: fig-ellipreduc
#| fig-cap: "A 2-d ellipse capturing most of the variability in a 3-d cloud of datapoints. Black points show original values on measured variables, blue lines show projections onto the new dimensions."
set.seed(33)
S = runif(3,.4,2)
f = runif(3,.7,.99)
R = f %*% t(f)
diag(R) = 1
R[1,3] <- R[2,3] <- R[3,1] <- R[3,2] <- .1
Sigma = diag(S)%*%R%*%diag(S)
Mean <- rep(0,3)
x <- MASS::mvrnorm(500, Mean, Sigma)


plot3d(x, box = FALSE, xlab="y1",ylab="y2",zlab="y3", alpha=.5)

xx = apply(x, 2, \(y) resid(lm(y ~ psych::principal(x,3,rotate="none")$scores[,3]))+
             rnorm(nrow(x),0,.001))
Sigma = cov(xx)
Mean = apply(xx,2,mean)
plot3d(ellipse3d(Sigma, centre = Mean, subdivisions = 50),
        col = "#A41AE4", alpha = 0.4, add = TRUE)
points3d(xx,add=T,col="#A41AE4")
for(i in 1:nrow(x)){
  segments3d(x = c(x[i,1],xx[i,1]),
              y = c(x[i,2],xx[i,2]),
              z = c(x[i,3],xx[i,3]),
              col="blue", lwd=.5, alpha=.5)

}
rglwidget()

```
:::
:::{.column width="50%"}
```{r}
#| echo: false
#| label: fig-ellipreduc2
#| fig-cap: "A 1-d line capturing most of the variability in a 3-d cloud of datapoints. Black points show original values on measured variables, blue lines show projections onto the new dimensions."
set.seed(33)
plot3d(x, box = FALSE, xlab="y1",ylab="y2",zlab="y3", alpha=.5)

# plot3d(abclines3d(0,0,0,a=principal(x,nfactors=3,rotate="none")$loadings[,1],
#              col = "#A41AE4"), lwd=2, add=TRUE)
xx = apply(x, 2, \(y) resid(lm(y ~ psych::principal(x,3,rotate="none")$scores[,2:3]))+
             rnorm(nrow(x),0,.0001))
# xx = apply(x, 2, \(y) resid(lm(y ~ rowMeans(x))))
points3d(xx,add=T,col="#A41AE4")
lines3d(xx,add=T,col="#A41AE4")
for(i in 1:nrow(x)){
  segments3d(x = c(x[i,1],xx[i,1]),
              y = c(x[i,2],xx[i,2]),
              z = c(x[i,3],xx[i,3]),
              col="blue", lwd=.5,lty="dotted",alpha=.5)

}

rglwidget()

```

:::
::::


The conceptual idea of these 'dimensions' that we're working with here is pretty abstract, and it only gets weirder to think about when we have more than 3 variables and we can no longer easily visualise it. 

The key thing we are doing here is:

1. Identify the dimensions that capture (co)variability across a set of measured variables
2. Retain a subset of those dimensions


Depending upon what the goal of our research is, the key questions are going to be:

1. how do we find the dimensions?
2. how many dimensions should we retain?
3. how does each dimension relate back to the original measured variables?  


# Key methods

TODO from here


## naive scale scores

e.g., rowMeans

Goal: pragmatic. i want 1 thing to use in some other setting (e.g., in a further analysis, or in a clinical setting)

1. how do we find the dimensions?
    - there's only one
2. how many dimensions should we retain?
    - there's only one
3. how does each dimension relate back to the original measured variables?  
    - equally


## PCA

Goal: pragmatic. i need fewer things for use in a subsequent analysis but don't want to lose too much variability.  

1. how do we find the dimensions?
    - set of orthogonal (i.e., perpendicular^[like all those that we've seen above]) dimensions that sequentially capture most variability
2. how many dimensions should we retain?
    - pragmatic: defined by either the number of dimensions we want, or the proportion of variability we want to retain
3. how does each dimension relate back to the original measured variables?  
    - examine correlations between dimensions and variables
    - somewhat irrelevant for 'pure' PCA, where we are agnostic/don't care about what the dimensions _are._  
    
    
## EFA

Goal: scale development.  i want to make a measure of construct X / I want to explore if/how many underlying dimensions there are to this measure.  

1. how do we find the dimensions?
    - estimation
    - models with different numbers of (possibly correlated) latent dimensions are compared
2. how many dimensions should we retain?
    - the model that best explains our observed relationships between variables
    - "best explains" = a theoretical question as much as a numerical one!  
3. how does each dimension relate back to the original measured variables?  
    - examine "loadings" between dimensions and variables
    
    
## CFA

Goal: theory testing. I have a theoretical model of how these variables relate to these constructs. I want to test how well that model is reflected in these data  

1. how do we find the dimensions?
    - we define them
2. how many dimensions should we retain?
    - irrelevant, the dimensions are pre-defined
3. how does each dimension relate back to the original measured variables?  
    - pre-defined
    


